---
title: "Editable Cells: the *_extra family"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Editable Cells: the *_extra family}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, message = FALSE, warning = FALSE)
```

reactablePlus gives you two ways to build an editable table, at two
different altitudes:

- **The `*_extra` cell factories** (this article) drop an editable input
  into a single [reactable](https://glin.github.io/reactable/) column.
  You keep full control of the `reactable()` call — its columns, theme,
  sorting, grouping, everything — and sprinkle in editable cells where
  you want them.
- **The config-driven module** (`table_config()` /
  `config_table_server()`, covered in
  [Getting Started](reactablePlus.html)) builds an *entire* editable
  table for you from a declarative spec, and layers on gating, selection,
  reset, mutual exclusion, fill-down, and three row modes.

Reach for `*_extra` when you already have a reactable and just need a
column or two to be editable. Graduate to the config module when the
table itself — its lifecycle, validation, and interactions — is the thing
you're building. The two share the same underlying machinery, so what you
learn here carries over.

## A first editable column

Every `*_extra` factory returns a **cell renderer** — a function suitable
for `reactable::colDef(cell = )`. Here is a read-only data frame with one
editable "comment" column:

```{r first}
library(shiny)
library(reactable)
library(reactablePlus)

tasks <- data.frame(
  id      = 1:3,
  task    = c("Draft proposal", "Review budget", "Send invites"),
  comment = c("", "", ""),
  stringsAsFactors = FALSE
)

ui <- fluidPage(
  useReactablePlus(),                 # 1. load the JS/CSS once, page-level
  reactableOutput("tbl"),
  verbatimTextOutput("out")
)

server <- function(input, output, session) {
  output$tbl <- renderReactable({
    bindExtrasOnRender(               # 3. attach the render hook
      reactable(
        tasks,
        columns = list(
          comment = colDef(
            html = TRUE,              # 2. opt the column into HTML cells
            cell = text_extra("comment", placeholder = "Add a note…")
          )
        )
      )
    )
  })

  output$out <- renderPrint({
    input$comment                     # a named list keyed by row index
  })
}

shinyApp(ui, server)
```

Three pieces of wiring are non-negotiable, and they map to the numbered
comments above:

1. **`useReactablePlus()`** at the page level. reactable renders cells
   with React and *drops* any `htmlDependency` returned from a cell
   function, so the supporting JS/CSS cannot ride along inside the cell —
   it has to be injected once at the top of the UI.
2. **`html = TRUE`** on the column's `colDef`, so reactable emits the
   factory's HTML instead of escaping it.
3. **`bindExtrasOnRender()`** wrapping the `reactable()` call. This
   attaches a render hook that seeds each column's value and keeps edits
   reconciled across repaints (more on that below).

Miss any one and the cells render but won't report values.

## The value-return contract

This is the single most important idea in the `*_extra` family.

**An entire column reports under one Shiny input.** The input is named by
the column id (prefixed by the module namespace; see
[Inside a module](#inside-a-module)). Its value is a **named list keyed by
the reactable row index**, in original data order:

```{r contract}
# For text_extra("comment") over the 3-row data frame above:
input$comment
#> list(`1` = "Call vendor", `2` = "", `3` = "Booked room")

# Flatten to a plain vector when you want one:
unlist(input$comment)
#>            1            2            3
#> "Call vendor"          ""  "Booked room"
```

Keying by **row index** rather than DOM position is deliberate: it keeps
the value mapped to the right record even after reactable sorts, filters,
or paginates. The list is fully seeded at render time, then updated
per-cell as the user types — so `input$comment` is populated immediately,
before any editing.

A few type-specific notes on what arrives in R:

- `numeric_extra` reports numbers; an empty cell reports as `NULL` (and so
  drops out of the named list).
- `date_extra` reports `"YYYY-MM-DD"` strings; empty cells report `NULL`.
- `checkbox_extra` and `toggle_extra` report logicals.
- Because JavaScript has no int/float distinction, whole numbers arrive as
  integers and fractional ones as doubles — test with `is.numeric()`, not
  `is.double()`.

## The factory family

All factories share the same first two arguments — `col_id` and `ns` —
and the same contract. They differ only in the control they render and the
type they coerce to.

```{r family}
# Free text
text_extra("comment", placeholder = "Optional…")

# Notes — text_extra plus the package's inline-input styling
notes_extra("notes", placeholder = "Context…")

# Numbers, with optional bounds and step
numeric_extra("qty", min = 0, max = 100, step = 1)

# Dates, with optional bounds (Date or "YYYY-MM-DD")
date_extra("due", min_date = "2024-01-01", max_date = "2030-12-31")

# Checkbox → TRUE / FALSE
checkbox_extra("done")

# On/off button → TRUE / FALSE, with custom labels
toggle_extra("billable", on_label = "Billable", off_label = "No charge")

# Dropdown — choices in any normalize_choices() format
dropdown_extra("status",
  choices = c("Open" = "open", "Blocked" = "blocked", "Done" = "done"),
  placeholder = "-- pick --"
)
```

`dropdown_extra()` accepts the same three choice formats as the rest of
the package — a plain vector, a named vector (names become labels), or a
list of `list(label, value)` — all normalized by `normalize_choices()`.
Supplying a `placeholder` adds a disabled empty option, so an unset cell
seeds as `""` rather than silently defaulting to the first option.

A small mixed table:

```{r mixed}
inventory <- data.frame(
  item     = c("Widgets", "Gadgets", "Gizmos"),
  qty      = c(0, 0, 0),
  status   = c("", "", ""),
  in_stock = c(FALSE, FALSE, FALSE),
  stringsAsFactors = FALSE
)

reactable(
  inventory,
  columns = list(
    qty = colDef(html = TRUE,
      cell = numeric_extra("qty", min = 0)),
    status = colDef(html = TRUE,
      cell = dropdown_extra("status",
        choices = c("In stock", "Backordered", "Discontinued"),
        placeholder = "-- status --")),
    in_stock = colDef(html = TRUE,
      cell = checkbox_extra("in_stock"))
  )
) |>
  bindExtrasOnRender()
```

## Collecting the edited data

Because each column is its own input, assembling a result data frame is a
matter of zipping the columns together by row index. A small helper keeps
empty cells aligned (remember: empty numeric/date cells drop out of the
list):

```{r collect}
# Read a column input into a length-n vector, filling gaps with `fill`.
collect_col <- function(col_input, n, fill = NA) {
  vapply(seq_len(n), function(i) {
    v <- col_input[[as.character(i)]]
    if (is.null(v)) fill else v
  }, FUN.VALUE = fill)
}

edited <- reactive({
  n <- nrow(inventory)
  data.frame(
    item     = inventory$item,
    qty      = collect_col(input$qty, n, NA_real_),
    status   = collect_col(input$status, n, NA_character_),
    in_stock = collect_col(input$in_stock, n, FALSE),
    stringsAsFactors = FALSE
  )
})
```

For the common all-text case you can lean on `unlist(input$comment)`
directly; the helper above is the robust version for mixed/optional
columns.

## Changing cells from the server

Edits flow client → server automatically. To push the other way — clear a
column, copy a value down, apply a computed default — use
`update_extra()`. It writes the authoritative value into the client-side
store *and* the live DOM, so the change survives reactable's repaints:

```{r update-extra}
# Clear every comment (3 rows), keyed by row index:
observeEvent(input$clear, {
  update_extra(session, "comment",
    values = setNames(as.list(rep("", 3)), as.character(1:3))
  )
})

# Copy row 1's status down to rows 2 and 3:
observeEvent(input$fill_down, {
  v <- input$status[["1"]]
  update_extra(session, "status",
    values = list(`2` = v, `3` = v)
  )
})
```

`values` is a named list keyed by the same row ids you see in the input
(here, the row index). Rows you omit are left untouched. Inside a module,
pass `ns = session$ns("")` so the message reaches the namespaced input.

Why a dedicated function instead of re-rendering? On every repaint the
store **wins**: it re-applies stored values to freshly painted cells so a
sort or filter never visually reverts an edit. A plain server-side
re-render would fight that reconcile; `update_extra()` updates the store
itself, so the new value sticks.

## Sorting and filtering: the one caveat

Edited values live in the JS store, **not** in reactable's data. reactable
sorts and filters by its original load-time values, so an editable column
would order rows by data the user can no longer see once they start
typing. The fix is simple:

```{r sortable}
status = colDef(
  html       = TRUE,
  sortable   = FALSE,    # don't sort by stale load-time values
  filterable = FALSE,
  cell       = dropdown_extra("status", choices = c("Open", "Done"))
)
```

Set `sortable = FALSE` (and `filterable = FALSE`) on `*_extra` columns,
and sort/filter on the table's read-only columns instead.

## Inside a module {#inside-a-module}

The `*_extra` factories are namespace-aware. In a Shiny module, pass
`ns = session$ns("")` so the column reports under the module's namespaced
input, and use the same prefix in `update_extra()`:

```{r module}
my_table_ui <- function(id) {
  ns <- NS(id)
  tagList(
    useReactablePlus(),
    reactableOutput(ns("tbl"))
  )
}

my_table_server <- function(id, df) {
  moduleServer(id, function(input, output, session) {
    output$tbl <- renderReactable({
      bindExtrasOnRender(reactable(df, columns = list(
        comment = colDef(html = TRUE,
          cell = text_extra("comment", ns = session$ns("")))
      )))
    })

    # input$comment is namespaced automatically; update_extra needs the ns:
    observeEvent(input$reset, {
      update_extra(session, "comment",
        values = setNames(as.list(rep("", nrow(df))),
                          as.character(seq_len(nrow(df)))),
        ns = session$ns("")
      )
    })
  })
}
```

`useReactablePlus()` is also safe to call once at the app level instead of
in every module UI — whichever fits your layout.

## v1 assumption: render all rows

The store seeds itself from the cells present at first paint, so the
current design assumes **pagination is off** (all rows rendered at
startup). With server-side pagination, rows not yet painted won't seed
until they appear. For editable `*_extra` tables, render the full set
(`defaultPageSize` large enough, or `pagination = FALSE`).

## When to graduate to `config_table`

The `*_extra` family is intentionally low-level: it edits cells and
reports values, and leaves everything else to you. If you find yourself
hand-rolling any of the following, the
[config-driven module](reactablePlus.html) already has it:

- conditional inputs that lock until other cells are set (**gating**),
- a column that clears another when set (**mutual exclusion**),
- a "copy this down to the rows below" affordance (**fill-down**),
- row **selection**, a **reset** button, or show/hide **gear toggles**,
- **dynamic** rows from a reactive data frame, or **appendable**
  user-managed rows,
- server-side **validation** and gate-enforced output.

Under the hood, the config module renders its primitive columns *and its
popover pickers* as these very same store-backed cells — so the contract
you learned here is exactly what `config_table_server()` reads from.

## See also

- [Getting Started with reactablePlus](reactablePlus.html) — the
  config-driven table module end to end.
- [Picker widgets and rich interactions](pickers-and-interactions.html) —
  typeahead search, multi-field popovers, mutual exclusion, and fill-down.
- `?text_extra`, `?update_extra`, `?bindExtrasOnRender` for the reference
  pages.
