---
title: "Using Config Table with reactablePlus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Config Table with reactablePlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```
This article walks through a practical use case for `config_table`:
a student grade entry form backed by a SQLite database. We define
the table declaratively with `table_config()` and `widget_col()`,
wire it into a Shiny app, and add persistence with a few helper
functions.

## The table

We want a table where a teacher can see each student-course pair,
pick a letter grade from a dropdown, and optionally add notes. Three
students, two courses, six rows.

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "code to create student roster"

library(shiny)
library(reactablePlus)

roster <- data.frame(
  row_id       = c("s1_c1", "s1_c2", "s2_c1", "s2_c2", "s3_c1", "s3_c2"),
  student_name = c("Amara Johnson", "Amara Johnson",
                   "Ben Martinez",  "Ben Martinez",
                   "Clara Chen",    "Clara Chen"),
  class_name   = c("Geometry", "Literature",
                   "Geometry", "Literature",
                   "Geometry", "Literature"),
  student_id   = c(1L, 1L, 2L, 2L, 3L, 3L),
  class_id     = c(1L, 2L, 1L, 2L, 1L, 2L),
  stringsAsFactors = FALSE
)
```
```{r print-roster}
#| eval: true
#| echo: true
library(dplyr)
glimpse(roster)
```

The `row_id` column uniquely identifies each row. `student_name` and
`class_name` are display-only. `student_id` and `class_id` travel
along silently — we'll need them later for saving.

Now define the table with `table_config()` and `widget_col()`.

```{r}
cfg <- table_config(
  row_keys   = roster$row_id,
  row_labels = paste(roster$student_name, "—", roster$class_name),

  columns = list(
    widget_col("grade", "dropdown", "Grade",
      options = list(
        choices = c("A+", "A", "A-",
                    "B+", "B", "B-",
                    "C+", "C", "C-",
                    "D+", "D", "D-",
                    "F")
      )
    ),
    widget_col("notes", "text", "Notes",
      min_width = 200,
      options = list(max_chars = 200, placeholder = "Optional notes...")
    )
  ),

  # Show the student-course labels as a badge column
  badge_col   = "label",
  badge_label = "Student / Course",

  show_reset = TRUE,

  to_output_fn = function(row_state, row_key) {
    data.frame(
      row_id = row_key,
      grade  = if (is.null(row_state$grade) || is.na(row_state$grade)) {
        NA_character_
      } else {
        as.character(row_state$grade)
      },
      notes = row_state$notes %||% "",
      stringsAsFactors = FALSE
    )
  },

  from_saved_fn = function(db_row, col_specs) {
    list(
      grade = if (!is.null(db_row$grade)) as.character(db_row$grade) else NA_character_,
      notes = if (!is.null(db_row$notes)) as.character(db_row$notes) else ""
    )
  }
)
```

The grade choices are a plain character vector — the simplest format.
`normalize_choices()` converts them internally so label and value are
the same string.

Wire it into a minimal app.

```{r}
ui <- fluidPage(
  titlePanel("Student Grade Entry"),
  config_table_ui("grades", cfg)
)

server <- function(input, output, session) {
  config_table_server("grades", cfg)
}

shinyApp(ui, server)
```

That's a working editable table with a reset button. Every row gets
a grade dropdown and a notes field, rendered inline in a reactable.
The module handles input wiring, state tracking, and data collection.

But the data disappears when the app restarts. Let's fix that.


## Adding a database

We'll use a single SQLite table to store grade entries. One row per
student-course pair, with an auto-incrementing ID.

```{r}
library(DBI)
library(RSQLite)

create_database <- function(db_path = "grades.sqlite") {
  con <- dbConnect(SQLite(), db_path)

  dbExecute(con, "
    CREATE TABLE IF NOT EXISTS student_grades (
      id         INTEGER PRIMARY KEY AUTOINCREMENT,
      student_id INTEGER NOT NULL,
      class_id   INTEGER NOT NULL,
      grade      TEXT,
      notes      TEXT,
      UNIQUE(student_id, class_id)
    )
  ")

  dbDisconnect(con)
  message("Database created at: ", db_path)
}

# Run once
create_database()
```

The `UNIQUE(student_id, class_id)` constraint means each student can
have at most one grade per course. We'll use this for upserts when
saving.


## Loading saved grades

When the app starts, we check if any grades were saved previously and
feed them back into the table via `data_r`.

```{r}
load_existing <- function(con, roster) {
  saved <- dbReadTable(con, "student_grades")
  if (nrow(saved) == 0) return(NULL)

  saved |>
    dplyr::mutate(row_id = paste0("s", student_id, "_c", class_id)) |>
    dplyr::select(row_id, grade, notes)
}
```

The `from_saved_fn` we defined in the config handles parsing each
saved row into the column state that `config_table_server` expects.


## Saving grades

When the teacher clicks Save, we read the module's `get_data()`
reactive — a data frame with `row_id`, `grade`, and `notes` — and
upsert each row into the database.

```{r}
save_grades <- function(con, current_data, roster) {
  if (is.null(current_data) || nrow(current_data) == 0) return()

  to_save <- current_data |>
    dplyr::left_join(
      roster |> dplyr::select(row_id, student_id, class_id),
      by = "row_id"
    ) |>
    dplyr::filter(!is.na(grade) & nchar(grade) > 0)

  if (nrow(to_save) == 0) return()

  purrr::walk(seq_len(nrow(to_save)), function(i) {
    row <- to_save[i, ]
    dbExecute(con, "
      INSERT INTO student_grades (student_id, class_id, grade, notes)
      VALUES (?, ?, ?, ?)
      ON CONFLICT(student_id, class_id)
      DO UPDATE SET grade = ?, notes = ?
    ", params = list(
      row$student_id, row$class_id, row$grade, row$notes,
      row$grade, row$notes
    ))
  })

  showNotification(
    paste("Saved", nrow(to_save), "grade entries."),
    type = "message"
  )
}
```


## The complete app

Everything wired together.

```{r}
library(shiny)
library(reactablePlus)
library(DBI)
library(RSQLite)
library(dplyr)

# ── Roster ────────────────────────────────────────────────────────────
roster <- data.frame(
  row_id       = c("s1_c1", "s1_c2", "s2_c1", "s2_c2", "s3_c1", "s3_c2"),
  student_name = c("Amara Johnson", "Amara Johnson",
                   "Ben Martinez",  "Ben Martinez",
                   "Clara Chen",    "Clara Chen"),
  class_name   = c("Geometry", "Literature",
                   "Geometry", "Literature",
                   "Geometry", "Literature"),
  student_id   = c(1L, 1L, 2L, 2L, 3L, 3L),
  class_id     = c(1L, 2L, 1L, 2L, 1L, 2L),
  stringsAsFactors = FALSE
)

# ── Config ────────────────────────────────────────────────────────────
cfg <- table_config(
  row_keys   = roster$row_id,
  row_labels = paste(roster$student_name, "\u2014", roster$class_name),

  columns = list(
    widget_col("grade", "dropdown", "Grade",
      options = list(
        choices = c("A+", "A", "A-", "B+", "B", "B-",
                    "C+", "C", "C-", "D+", "D", "D-", "F")
      )
    ),
    widget_col("notes", "text", "Notes",
      min_width = 200,
      options = list(max_chars = 200, placeholder = "Optional notes...")
    )
  ),

  badge_col   = "label",
  badge_label = "Student / Course",
  show_reset  = TRUE,

  to_output_fn = function(row_state, row_key) {
    data.frame(
      row_id = row_key,
      grade  = if (is.null(row_state$grade) || is.na(row_state$grade)) {
        NA_character_
      } else {
        as.character(row_state$grade)
      },
      notes = row_state$notes %||% "",
      stringsAsFactors = FALSE
    )
  },

  from_saved_fn = function(db_row, col_specs) {
    list(
      grade = if (!is.null(db_row$grade)) as.character(db_row$grade) else NA_character_,
      notes = if (!is.null(db_row$notes)) as.character(db_row$notes) else ""
    )
  }
)

# ── Helpers ───────────────────────────────────────────────────────────
load_existing <- function(con, roster) {
  saved <- dbReadTable(con, "student_grades")
  if (nrow(saved) == 0) return(NULL)

  saved |>
    mutate(row_id = paste0("s", student_id, "_c", class_id)) |>
    select(row_id, grade, notes)
}

save_grades <- function(con, current_data, roster) {
  if (is.null(current_data) || nrow(current_data) == 0) return()

  to_save <- current_data |>
    left_join(
      roster |> select(row_id, student_id, class_id),
      by = "row_id"
    ) |>
    filter(!is.na(grade) & nchar(grade) > 0)

  if (nrow(to_save) == 0) return()

  purrr::walk(seq_len(nrow(to_save)), function(i) {
    row <- to_save[i, ]
    dbExecute(con, "
      INSERT INTO student_grades (student_id, class_id, grade, notes)
      VALUES (?, ?, ?, ?)
      ON CONFLICT(student_id, class_id)
      DO UPDATE SET grade = ?, notes = ?
    ", params = list(
      row$student_id, row$class_id, row$grade, row$notes,
      row$grade, row$notes
    ))
  })

  showNotification(
    paste("Saved", nrow(to_save), "grade entries."),
    type = "message"
  )
}

# ── App ───────────────────────────────────────────────────────────────
ui <- fluidPage(
  titlePanel("Student Grade Entry"),
  config_table_ui("grades", cfg),
  br(),
  actionButton("save", "Save Grades", class = "btn-primary")
)

server <- function(input, output, session) {
  con <- dbConnect(SQLite(), "grades.sqlite")
  onStop(function() dbDisconnect(con))

  existing_data <- load_existing(con, roster)

  result <- config_table_server("grades", cfg,
    data_r = reactive(existing_data)
  )

  observeEvent(input$save, {
    save_grades(con, result$get_data(), roster)
  })
}

shinyApp(ui, server)
```

Run `create_database()` once, then launch the app. Pick grades, add
notes, click Save. Restart the app — your entries are still there.


## What the module handles for you

This six-row table has 12 inputs (6 dropdowns + 6 text fields). A
table with 30 students and 5 courses would have 300. Without
`config_table_server()`, you'd build and track each one manually,
synchronize them with the reactable display, and collect all the
values back into a data frame yourself. The module reduces that to
a `table_config()`, a couple of `widget_col()` calls, and a reactive.

The pattern shown here — load from a database, edit in a table,
save back — generalizes to any domain: inventory management, survey
responses, data annotation, scheduling, or anywhere you need
structured inline editing in Shiny.
