---
title: "editable_table"
execute:
  eval: false
format:
  html:
    code-copy: true
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

The `editable_table_server()` and `editable_table_ui()` module functions are designed to assist with uploading data user entry data to databases. 


The `editable_table_ui()` function returns a list of shiny tags: 

*  __table__ - A reactable table for the user to interact with
*  __reset_button__ - A reset button
*  __data_display__ - A table showing the output of the user selections

The `data_display` table can be used as a diagnostic for troubleshooting the module, or sending the user feedback.


This article walks through a practical use case for the raw-spec
editable table module: a student grade entry form backed by a SQLite
database. We'll start with the table itself, then add persistence.


The `editable_table_server()` returns a named list with two reactives:

*  __current_data__ - This is a tibble of the data the user has selected. It includes an id_col along with whatever values the user has selected.
*  __selected_ids__ - This is a vector of id_col values that have been selected by the user. When a row is NOT selectable, it returns NULL. This return item can be used when you only need a list of the values selected.

## The table

We want a table where a teacher can see each student-course pair,
pick a letter grade from a dropdown, and optionally add notes. Three
students, two courses, six rows.

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "code to create student roster"

library(shiny)
library(reactablePlus)

roster <- data.frame(
  row_id       = c("s1_c1", "s1_c2", "s2_c1", "s2_c2", "s3_c1", "s3_c2"),
  student_name = c("Amara Johnson", "Amara Johnson",
                   "Ben Martinez",  "Ben Martinez",
                   "Clara Chen",    "Clara Chen"),
  class_name   = c("Geometry", "Literature",
                   "Geometry", "Literature",
                   "Geometry", "Literature"),
  student_id   = c(1L, 1L, 2L, 2L, 3L, 3L),
  class_id     = c(1L, 2L, 1L, 2L, 1L, 2L)
)
```
```{r print-roster}
#| eval: true
#| echo: true
library(dplyr)
glimpse(roster)
```



The `row_id` column uniquely identifies each row. `student_name` and
`class_name` are display-only. `student_id` and `class_id` travel
along silently — we'll need them later for saving.

Now define the table layout with the constructor functions.

```{r}
rows <- row_def(
  id_col = "row_id",
  display_cols = list(
    display_col("student_name", "Student"),
    display_col("class_name", "Course")
  )
)

cols <- list(
  input_col("grade", "Grade", "dropdown",
            choices = c("A+", "A", "A-",
                        "B+", "B", "B-",
                        "C+", "C", "C-",
                        "D+", "D", "D-",
                        "F")),
  input_col("notes", "Notes", "text", max_chars = 200)
)
```

The grade choices are a plain character vector — the simplest format.
Label and value are the same string. The module normalizes them
internally via `normalize_choices()`.

Wire it into a minimal app.

```{r}
ui <- fluidPage(
  titlePanel("Student Grade Entry"),
  editable_table_ui("grades")
)

server <- function(input, output, session) {
  editable_table_server(
    "grades",
    data_r   = reactive(roster),
    row_spec = rows,
    col_spec = cols
  )
}

shinyApp(ui, server)
```

That's a working editable table in about 30 lines. Every row gets a
grade dropdown and a notes field, rendered inline in a reactable. The
module handles all the input wiring, state tracking, and data
collection.

But the data disappears when the app restarts. Let's fix that.


## Adding a database

We'll use a single SQLite table to store grade entries. One row per
student-course pair, with an auto-incrementing ID.

```{r}
library(DBI)
library(RSQLite)

create_database <- function(db_path = "grades.sqlite") {
  con <- dbConnect(SQLite(), db_path)

  dbExecute(con, "
    CREATE TABLE IF NOT EXISTS student_grades (
      id         INTEGER PRIMARY KEY AUTOINCREMENT,
      student_id INTEGER NOT NULL,
      class_id   INTEGER NOT NULL,
      grade      TEXT,
      notes      TEXT,
      UNIQUE(student_id, class_id)
    )
  ")

  dbDisconnect(con)
  message("Database created at: ", db_path)
}

# Run once
create_database()
```

The `UNIQUE(student_id, class_id)` constraint means each student can
have at most one grade per course. We'll use this for upserts when
saving.


## Loading saved grades

When the app starts, we need to check if any grades were saved
previously and feed them back into the table.

```{r}
load_existing <- function(con, roster) {
  saved <- dbReadTable(con, "student_grades")
  if (nrow(saved) == 0) return(NULL)

  saved |>
    dplyr::mutate(row_id = paste0("s", student_id, "_c", class_id)) |>
    dplyr::select(row_id, grade, notes)
}
```

The module's `existing_data_r` parameter accepts a data frame with
the same `row_id` values and one column per editable field. The
module merges these into the initial table state automatically via
`merge_existing_data()`.


## Saving grades

When the teacher clicks Save, we read the module's `current_data`
reactive — a tibble with `row_id`, `grade`, and `notes` — and upsert
each row into the database.

```{r}
save_grades <- function(con, current_data, roster) {
  if (is.null(current_data) || nrow(current_data) == 0) return()

  to_save <- current_data |>
    dplyr::left_join(
      roster |> dplyr::select(row_id, student_id, class_id),
      by = "row_id"
    ) |>
    dplyr::filter(!is.na(grade) & nchar(grade) > 0)

  if (nrow(to_save) == 0) return()

  purrr::walk(seq_len(nrow(to_save)), function(i) {
    row <- to_save[i, ]
    dbExecute(con, "
      INSERT INTO student_grades (student_id, class_id, grade, notes)
      VALUES (?, ?, ?, ?)
      ON CONFLICT(student_id, class_id)
      DO UPDATE SET grade = ?, notes = ?
    ", params = list(
      row$student_id, row$class_id, row$grade, row$notes,
      row$grade, row$notes
    ))
  })

  message("Saved ", nrow(to_save), " grade entries.")
}
```

The `ON CONFLICT ... DO UPDATE` clause handles both inserts and
updates in a single statement. If a student-course pair already
exists, the grade and notes are overwritten.


## The complete app

Everything wired together.

```{r}
library(shiny)
library(reactablePlus)
library(DBI)
library(RSQLite)
library(dplyr)

# ── Roster (in a real app, this might come from the database too) ────
roster <- data.frame(
  row_id       = c("s1_c1", "s1_c2", "s2_c1", "s2_c2", "s3_c1", "s3_c2"),
  student_name = c("Amara Johnson", "Amara Johnson",
                   "Ben Martinez",  "Ben Martinez",
                   "Clara Chen",    "Clara Chen"),
  class_name   = c("Geometry", "Literature",
                   "Geometry", "Literature",
                   "Geometry", "Literature"),
  student_id   = c(1L, 1L, 2L, 2L, 3L, 3L),
  class_id     = c(1L, 2L, 1L, 2L, 1L, 2L)
)

# ── Table layout ────────────────────────────────────────────────────
rows <- row_def(
  id_col = "row_id",
  display_cols = list(
    display_col("student_name", "Student"),
    display_col("class_name", "Course")
  )
)

cols <- list(
  input_col("grade", "Grade", "dropdown",
            choices = c("A+", "A", "A-",
                        "B+", "B", "B-",
                        "C+", "C", "C-",
                        "D+", "D", "D-",
                        "F")),
  input_col("notes", "Notes", "text", max_chars = 200)
)

# ── Helpers ─────────────────────────────────────────────────────────
load_existing <- function(con, roster) {
  saved <- dbReadTable(con, "student_grades")
  if (nrow(saved) == 0) return(NULL)

  saved |>
    mutate(row_id = paste0("s", student_id, "_c", class_id)) |>
    select(row_id, grade, notes)
}

save_grades <- function(con, current_data, roster) {
  if (is.null(current_data) || nrow(current_data) == 0) return()

  to_save <- current_data |>
    left_join(
      roster |> select(row_id, student_id, class_id),
      by = "row_id"
    ) |>
    filter(!is.na(grade) & nchar(grade) > 0)

  if (nrow(to_save) == 0) return()

  purrr::walk(seq_len(nrow(to_save)), function(i) {
    row <- to_save[i, ]
    dbExecute(con, "
      INSERT INTO student_grades (student_id, class_id, grade, notes)
      VALUES (?, ?, ?, ?)
      ON CONFLICT(student_id, class_id)
      DO UPDATE SET grade = ?, notes = ?
    ", params = list(
      row$student_id, row$class_id, row$grade, row$notes,
      row$grade, row$notes
    ))
  })

  showNotification(
    paste("Saved", nrow(to_save), "grade entries."),
    type = "message"
  )
}

# ── App ─────────────────────────────────────────────────────────────
ui <- fluidPage(
  titlePanel("Student Grade Entry"),
  editable_table_ui("grades"),
  br(),
  actionButton("save", "Save Grades", class = "btn-primary")
)

server <- function(input, output, session) {
  con <- dbConnect(SQLite(), "grades.sqlite")
  onStop(function() dbDisconnect(con))

  existing <- load_existing(con, roster)

  result <- editable_table_server(
    "grades",
    data_r          = reactive(roster),
    row_spec        = rows,
    col_spec        = cols,
    existing_data_r = reactive(existing)
  )

  observeEvent(input$save, {
    save_grades(con, result$current_data(), roster)
  })
}

shinyApp(ui, server)
```

Run `create_database()` once, then launch the app. Pick grades, add
notes, click Save. Restart the app — your entries are still there.


## What the module handles for you

This six-row table has 12 inputs (6 dropdowns + 6 text fields). A
table with 30 students and 5 courses would have 300. Without
`editable_table_server()`, you'd build and track each one manually,
synchronize them with the reactable display, and collect all the
values back into a data frame yourself. The module reduces that to
a `row_def()`, a couple of `input_col()` calls, and a reactive.

The pattern shown here — load from a database, edit in a table,
save back — generalizes to any domain: inventory management, survey
responses, data annotation, scheduling, or anywhere you need
structured inline editing in Shiny.
