---
title: "Building a Grade Entry System with editable_table"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article walks through a realistic use case for the raw-spec
editable table module: a student grade entry system backed by a SQLite
database. By the end, you'll have a Shiny app where a teacher can
select grades from a dropdown, add notes, and save everything back
to the database — with the module handling all the UI complexity.

## Why use the editable table module?

Building an editable table in Shiny from scratch means wiring up
individual inputs, tracking their state, synchronizing with reactable,
handling resets, and collecting values back into a data frame. For a
table with 10 students and 4 columns, that's 40 inputs to manage.

`editable_table_server()` handles all of that declaratively. You
describe the shape of the table — which columns are read-only, which
are editable, and what input type each one uses — and the module
handles rendering, state management, and data collection. Your app
code focuses on what matters: loading data and saving results.

## Setting up the database

We'll use a SQLite database with four tables: grades (the lookup table
for letter grades and GPA values), classes, students, and
student_grades (the junction table where entries get saved).

```{r}
library(DBI)
library(RSQLite)

create_database <- function(db_path = "grades.sqlite") {
  con <- dbConnect(SQLite(), db_path)

  # ── Grades lookup table ──────────────────────────────────────────────
  dbExecute(con, "
    CREATE TABLE IF NOT EXISTS grades (
      id           INTEGER PRIMARY KEY,
      letter_grade TEXT    NOT NULL,
      gpa_value    REAL    NOT NULL
    )
  ")

  grade_data <- data.frame(
    id           = 1:13,
    letter_grade = c("A+", "A", "A-",
                     "B+", "B", "B-",
                     "C+", "C", "C-",
                     "D+", "D", "D-",
                     "F"),
    gpa_value    = c(4.333, 4.0, 3.667,
                     3.333, 3.0, 2.667,
                     2.333, 2.0, 1.667,
                     1.333, 1.0, 0.667,
                     0.0)
  )

  dbExecute(con, "DELETE FROM grades")
  dbWriteTable(con, "grades", grade_data, append = TRUE)

  # ── Classes ──────────────────────────────────────────────────────────
  dbExecute(con, "
    CREATE TABLE IF NOT EXISTS classes (
      id         INTEGER PRIMARY KEY,
      class_name TEXT    NOT NULL
    )
  ")

  class_data <- data.frame(
    id         = 1:4,
    class_name = c("Geometry", "Literature", "Philosophy", "Science")
  )

  dbExecute(con, "DELETE FROM classes")
  dbWriteTable(con, "classes", class_data, append = TRUE)

  # ── Students ─────────────────────────────────────────────────────────
  dbExecute(con, "
    CREATE TABLE IF NOT EXISTS students (
      id         INTEGER PRIMARY KEY,
      first_name TEXT    NOT NULL,
      last_name  TEXT    NOT NULL
    )
  ")

  student_data <- data.frame(
    id         = 1:10,
    first_name = c("Amara", "Ben", "Clara", "David", "Elena",
                   "Felix", "Grace", "Henry", "Iris", "James"),
    last_name  = c("Johnson", "Martinez", "Chen", "Williams", "Petrov",
                   "Okafor", "Kim", "Dubois", "Nakamura", "Thompson")
  )

  dbExecute(con, "DELETE FROM students")
  dbWriteTable(con, "students", student_data, append = TRUE)

  # ── Student grades (junction table) ──────────────────────────────────
  dbExecute(con, "
    CREATE TABLE IF NOT EXISTS student_grades (
      id         INTEGER PRIMARY KEY AUTOINCREMENT,
      student_id INTEGER NOT NULL,
      class_id   INTEGER NOT NULL,
      grade_id   INTEGER,
      notes      TEXT,
      FOREIGN KEY (student_id) REFERENCES students(id),
      FOREIGN KEY (class_id)   REFERENCES classes(id),
      FOREIGN KEY (grade_id)   REFERENCES grades(id),
      UNIQUE(student_id, class_id)
    )
  ")

  dbDisconnect(con)
  message("Database created at: ", db_path)
}

# Run once to set up the database
create_database()
```

## Loading data for the table

The editable table module needs a data frame with at least an ID column
and any display columns. We'll query a cross-join of students and
classes so every student gets a row for every course.

```{r}
library(dplyr)

load_roster <- function(con) {
  students <- dbReadTable(con, "students")
  classes  <- dbReadTable(con, "classes")

  tidyr::crossing(
    students |> select(student_id = id, first_name, last_name),
    classes  |> select(class_id = id, class_name)
  ) |>
    mutate(
      row_id       = paste0("s", student_id, "_c", class_id),
      student_name = paste(first_name, last_name)
    ) |>
    select(row_id, student_name, class_name, student_id, class_id) |>
    arrange(last_name, first_name, class_name)
}
```

We also need to load any previously saved grades, so returning users
see their earlier entries.

```{r}
load_existing_grades <- function(con) {
  saved <- dbReadTable(con, "student_grades")
  if (nrow(saved) == 0) return(NULL)

  grades <- dbReadTable(con, "grades")

  saved |>
    left_join(grades, by = c("grade_id" = "id")) |>
    mutate(
      row_id = paste0("s", student_id, "_c", class_id),
      grade  = letter_grade
    ) |>
    select(row_id, grade, notes)
}
```


## Building the grade choices

The grade dropdown should display letter grades (A+, A, A-, ...) as
both the label and the value. We pull them straight from the database.

```{r}
load_grade_choices <- function(con) {
  grades <- dbReadTable(con, "grades")
  grades$letter_grade
}
```

Since the labels and values are the same, we can pass a plain character
vector to `input_col()` — `normalize_choices()` handles the rest.


## Defining the table layout

Now we wire everything together with the constructor functions. The
display columns show the student name and course. The editable columns
are the grade dropdown and a notes text field.

```{r}
library(reactablePlus)

rows <- row_def(
  id_col = "row_id",
  display_cols = list(
    display_col("student_name", "Student"),
    display_col("class_name", "Course")
  )
)

build_col_spec <- function(grade_choices) {
  list(
    input_col("grade", "Grade", "dropdown", choices = grade_choices),
    input_col("notes", "Notes", "text", max_chars = 500)
  )
}
```

We wrap the column spec in a function because the grade choices come
from the database and aren't available until runtime.


## Saving results

When the user clicks Save, we collect the module's `current_data`
reactive and upsert it into the `student_grades` table. The module
returns a tibble with the `row_id` and one column per `input_col`.

```{r}
save_grades <- function(con, current_data, roster) {
  if (is.null(current_data) || nrow(current_data) == 0) return()

  grades_lookup <- dbReadTable(con, "grades")

  to_save <- current_data |>
    left_join(
      roster |> select(row_id, student_id, class_id),
      by = "row_id"
    ) |>
    left_join(
      grades_lookup |> select(grade_id = id, letter_grade),
      by = c("grade" = "letter_grade")
    ) |>
    filter(!is.na(grade_id)) |>
    select(student_id, class_id, grade_id, notes)

  if (nrow(to_save) == 0) return()

  # Upsert: replace existing entries, insert new ones
  purrr::walk(seq_len(nrow(to_save)), function(i) {
    row <- to_save[i, ]
    dbExecute(con, "
      INSERT INTO student_grades (student_id, class_id, grade_id, notes)
      VALUES (?, ?, ?, ?)
      ON CONFLICT(student_id, class_id)
      DO UPDATE SET grade_id = ?, notes = ?
    ", params = list(
      row$student_id, row$class_id, row$grade_id, row$notes,
      row$grade_id, row$notes
    ))
  })

  message("Saved ", nrow(to_save), " grade entries.")
}
```


## The complete app

Here's everything wired together into a working Shiny application.

```{r}
library(shiny)
library(reactablePlus)
library(DBI)
library(RSQLite)
library(dplyr)

ui <- fluidPage(
  titlePanel("Student Grade Entry"),
  editable_table_ui("grades"),
  br(),
  actionButton("save", "Save Grades", class = "btn-primary")
)

server <- function(input, output, session) {
  # ── Database connection ────────────────────────────────────────────
  con <- dbConnect(SQLite(), "grades.sqlite")
  onStop(function() dbDisconnect(con))

  # ── Load reference data ────────────────────────────────────────────
  roster        <- load_roster(con)
  grade_choices <- load_grade_choices(con)
  existing      <- load_existing_grades(con)

  # ── Define table layout ────────────────────────────────────────────
  rows <- row_def(
    id_col = "row_id",
    display_cols = list(
      display_col("student_name", "Student"),
      display_col("class_name", "Course")
    )
  )

  cols <- build_col_spec(grade_choices)

  # ── Render the editable table ──────────────────────────────────────
  result <- editable_table_server(
    "grades",
    data_r          = reactive(roster),
    row_spec        = rows,
    col_spec        = cols,
    existing_data_r = reactive(existing)
  )

  # ── Save on button click ───────────────────────────────────────────
  observeEvent(input$save, {
    save_grades(con, result$current_data(), roster)
    showNotification("Grades saved!", type = "message")
  })
}

shinyApp(ui, server)
```

When you run this app, you'll see a table with 40 rows (10 students ×
4 courses). Each row has a grade dropdown populated from the database
and a notes field. Click Save to write everything back to SQLite.

Restarting the app loads previous entries via `existing_data_r`,
so the teacher picks up right where they left off.


## What the module handles for you

Without `editable_table_server()`, building this app would mean
manually creating 40 dropdown inputs and 40 text inputs, tracking
which ones changed, handling the Reset button, collecting all 80
values back into a data frame, and keeping the reactable display
in sync. The module reduces that to a few lines of configuration.

The pattern shown here — load from a database, edit in a table,
save back — generalizes to any domain: inventory tracking, survey
entry, scheduling, data annotation, or anywhere you need structured
inline editing in Shiny.
