---
title: "Picker Widgets and Rich Interactions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Picker Widgets and Rich Interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, message = FALSE, warning = FALSE)
```

Some cell edits don't fit in a text box. You want to search a database of
thousands of options, fill out a small multi-field form, or capture a
structured rating — all without leaving the row. reactablePlus ships a
family of **popover picker widgets** for exactly this, and the
config-driven table module wires them together with three table-level
interactions: **mutual exclusion**, **fill-down**, and **gear toggles**.

This article covers both halves: the pickers themselves, and the
interactions that make a table of them feel like an application. If you
haven't met the config module yet, skim
[Getting Started](reactablePlus.html) first — and to see everything here
running at once, launch the bundled demo:

```{r demo}
reactablePlus::runExample("school-history")
```

## The picker family

Each picker is a standalone Shiny input backed by a custom JS binding and
a shared popover. They render a compact trigger in the cell; clicking it
opens a popover; the chosen value flows back to `input$<id>`.

| Widget | `widget_col` type | Reports |
|---|---|---|
| `searchPickerInput()` | `"search_picker"` | the picked record (a list), or `NULL` |
| `attendancePickerInput()` | `"attendance_picker"` | a list keyed by section, or `NULL` |
| `homeschoolPickerInput()` | `"homeschool_picker"` | `list(by, by_other, curriculum, notes)`, or `NULL` when off |
| `notesInput()` | `"notes_input"` | a string |
| `gearPopoverInput()` | — (table toolbar) | a named list of toggle states |

All of them share one rule: because reactable drops cell-level HTML
dependencies, the supporting JS/CSS must be loaded once at the page level
with **`useReactablePlus()`**. Inside the config module that call is
automatic; standalone, you make it yourself.

### Standalone use

Any picker works as a plain Shiny input, table or no table:

```{r standalone}
library(shiny)
library(reactablePlus)

ui <- fluidPage(
  useReactablePlus(),
  homeschoolPickerInput("hs", grade_label = "Grade 3"),
  attendancePickerInput("att", grade_label = "Grade 3"),
  notesInput("note", placeholder = "Anything else?"),
  verbatimTextOutput("out")
)

server <- function(input, output, session) {
  output$out <- renderPrint({
    list(homeschool = input$hs, attendance = input$att, note = input$note)
  })

  # Push a value from the server:
  observeEvent(input$preset, {
    updateHomeschoolPickerInput(session, "hs",
      value = list(by = "Mother", curriculum = "Sonlight"))
  })
}

shinyApp(ui, server)
```

Each constructor has a matching `update*()` function
(`updateSearchPickerInput()`, `updateAttendancePickerInput()`,
`updateHomeschoolPickerInput()`, `updateNotesInput()`,
`updateGearPopoverInput()`) for server-driven changes. The `grade_label`
argument is just an accessible label for the trigger — the widgets are
domain-agnostic; "grade" is a holdover from the package's origin and can
read as anything.

## Typeahead search

`searchPickerInput()` is the heavyweight of the family: a debounced
typeahead over a backend you supply. You never ship the option list to the
browser — the widget sends queries to the server, your function answers,
and the picked row comes back as the value.

The server side is one call, `useTypeaheadSearch()`, given a function
`(query, limit) -> data.frame`:

```{r typeahead-standalone}
ui <- fluidPage(
  useReactablePlus(),
  searchPickerInput("school", trigger_label = "+ Pick school"),
  verbatimTextOutput("picked")
)

server <- function(input, output, session) {
  schools <- data.frame(
    id    = c("S1", "S2", "S3"),
    name  = c("Lincoln Elementary", "Washington Middle", "Jefferson High"),
    city  = c("Springfield", "Springfield", "Capital City"),
    stringsAsFactors = FALSE
  )

  useTypeaheadSearch(input, session, search_fn = function(query, limit) {
    hits <- grepl(tolower(query), tolower(schools$name))
    head(schools[hits, , drop = FALSE], limit)
  })

  output$picked <- renderPrint(input$school)   # the chosen row, as a list
}
```

Each returned data-frame row becomes a candidate; the row the user picks
becomes the widget's value — a list with that row's columns
(`input$school$name`, `input$school$id`, …). Include whatever fields your
app needs downstream; you'll see in [Fill-down](#fill-down) that extra
columns like a valid range can ride along on the value.

### Search inside the config module

In a `table_config()`, you don't call `useTypeaheadSearch()` yourself.
Name the search column with `search_fn_col`, and hand the search function
to `config_table_server()`:

```{r typeahead-config}
cfg <- table_config(
  row_keys   = c("r1", "r2", "r3"),
  row_labels = c("Slot 1", "Slot 2", "Slot 3"),
  columns = list(
    widget_col("school", "search_picker", "School", min_width = 300)
  ),
  search_fn_col = "school"
)

server <- function(input, output, session) {
  config_table_server("tbl", cfg, search_fn = function(query, limit) {
    head(schools[grepl(tolower(query), tolower(schools$name)), ], limit)
  })
}
```

The module routes results to whichever picker asked, across every row.

## Pickers in a table

Putting it together: a school-history record where each grade row can hold
a searched school, an attendance summary, a homeschool form, and free
notes. This mirrors the bundled `school-history` demo.

```{r table}
library(shiny)
library(reactablePlus)
library(reactable)

cfg <- table_config(
  row_keys   = gradeChoices(),                 # PK, K, 01 … 12
  row_labels = names(gradeChoices()),

  columns = list(
    widget_col("school", "search_picker", "School",
      min_width = 300, triggers_rerender = TRUE,
      options = list(show_fill_down = TRUE)),
    widget_col("homeschool", "homeschool_picker", "Homeschool",
      width = 200, triggers_rerender = TRUE,
      gear_toggle = "showHomeschool"),
    widget_col("attendance", "attendance_picker", "Attendance",
      width = 220),
    widget_col("notes", "notes_input", "Notes", min_width = 150)
  ),

  search_fn_col = "school",
  badge_col = "grade", badge_label = "Grade"
)

server <- function(input, output, session) {
  result <- config_table_server("history", cfg, search_fn = search_schools)
  # result$get_data() is your reactive data frame.
}
```

`triggers_rerender = TRUE` tells the module that changing this column may
affect others (here, because school and homeschool exclude each other, and
because the gate/visibility state can shift), so the table re-renders on
change. Now the three interactions.

## Mutual exclusion

Two columns that can't both be set: picking a school should clear the
homeschool entry for that row, and vice versa. Declare it once under
`interactions$mutual_exclusion` — a list of rules, each saying *when this
column turns on, clear that one* and what placeholder to show in the
cleared cell:

```{r mutual-exclusion}
interactions = list(
  mutual_exclusion = list(
    list(
      when_on = "homeschool", clears = "school",
      display = '<span style="opacity:.5; font-style:italic;">homeschooled</span>'
    ),
    list(when_on = "school", clears = "homeschool", display = "")
  )
)
```

When a rule fires, the module clears the target in server state, swaps the
cleared cell to show its `display` HTML, and — because both columns are
`triggers_rerender` — refreshes the row. The cleared widget reports its
empty value (`NULL`) on the next `get_data()`. The display swap happens
in place, without a full table re-render, so it stays snappy.

A natural companion is `row_class_fn`, to restyle the whole row when a
column is active:

```{r row-class}
row_class_fn = function(row_key, row_state) {
  if (!is.null(row_state$homeschool)) "is-homeschool" else NULL
}
```

## Fill-down {#fill-down}

A student usually attends the same school for several grades in a row.
Fill-down lets the user set one cell and cascade it down to the empty rows
below — bounded by a range check. Enable the affordance on the picker
(`show_fill_down = TRUE`) and describe the cascade under
`interactions$fill_down`:

```{r fill-down}
interactions = list(
  fill_down = list(
    column = "school",
    range_check_fn = function(row_key, value) {
      # `value` is the picked record; cascade only into grades the
      # school actually serves.
      gradeInRange(row_key, value$low_grade, value$high_grade)
    }
  )
)
```

Clicking the picker's down-arrow sends a fill-down request; the module
reads the source value from server state and writes it into every
**empty**, **in-range** row beneath the source — skipping rows already
filled, rows where a mutually-exclusive column is active, and rows your
`range_check_fn` rejects. This is why it pays to carry extra fields
(`low_grade`, `high_grade`) on the search result: the range check reads
them straight off the value.

`fill_down` takes an optional `input_name` (the Shiny input the widget's
link targets; defaults to `paste0(column, "_fill_down")`) — you rarely
need to set it.

## Gear toggles

A gear popover in the toolbar holds per-table view options — show/hide
columns, display modes, density. Define them as a named list of
`list(label, value)` via `gear_toggles`, then bind a column's visibility
to a toggle key with `widget_col(gear_toggle = ...)`:

```{r gear}
cfg <- table_config(
  row_keys   = gradeChoices(),
  row_labels = names(gradeChoices()),

  columns = list(
    widget_col("school", "search_picker", "School"),
    # Hidden until the "showHomeschool" toggle is on:
    widget_col("homeschool", "homeschool_picker", "Homeschool",
      gear_toggle = "showHomeschool")
  ),

  gear_toggles = list(
    showNCESId     = list(label = "Show NCES ID",  value = TRUE),
    showHomeschool = list(label = "Show Homeschool", value = FALSE),
    compactRows    = list(label = "Compact rows",  value = FALSE)
  )
)
```

Toggling `showHomeschool` reveals or hides the homeschool column in place.
Other toggles (like `showNCESId` or `compactRows`) are surfaced to the
module's render settings, so picker widgets and CSS can read them — for
example, the search picker shows or hides each result's identifier based
on `showNCESId`.

Standalone, the same popover is `gearPopoverInput()`, reporting a named
list of toggle states you can react to yourself:

```{r gear-standalone}
gearPopoverInput("view_opts", toggles = list(
  dense  = list(label = "Dense rows", value = FALSE),
  totals = list(label = "Show totals", value = TRUE)
))
# server: observe(input$view_opts$dense), update via updateGearPopoverInput()
```

## Putting it together

These interactions are declarative and composable — mutual exclusion,
fill-down, gating ([Getting Started](reactablePlus.html#gating)), gear
toggles, selection, and reset all coexist in one `table_config()`. The
`school-history` demo combines every one of them over a PreK–12 grade
ladder; reading its source is the fastest way to see the wiring in a real
app:

```{r source}
file.edit(system.file("examples/school-history/app.R",
                      package = "reactablePlus"))
```

## See also

- [Getting Started with reactablePlus](reactablePlus.html) — the table
  module, gating, selection, dynamic and appendable rows.
- [Editable Cells: the *_extra family](editable-cells.html) — the
  lower-level cell factories the module is built on.
- `?searchPickerInput`, `?useTypeaheadSearch`, `?table_config`,
  `?widget_col` for the reference pages.
