---
title: "Getting Started with reactablePlus"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

reactablePlus extends [reactable](https://glin.github.io/reactable/) with
inline editing widgets and two Shiny module layers for building editable
tables. This article walks through the key concepts.

## Installation

```{r}
# install.packages("pak")
pak::pak("matthewcurrier/reactablePlus")
```

## Two ways to build editable tables

reactablePlus offers two abstraction levels. Pick the one that fits your use
case.

### 1. Raw-spec module (maximum flexibility)

The lower-level module accepts declarative row and column definitions.
You control every column type (dropdown, numeric, checkbox, toggle, date,
text) and can wire gating, selection, and reset behavior.

Use constructor functions — `row_def()`, `display_col()`, and `input_col()`
— to build validated specifications. They catch configuration errors at
creation time instead of at render time.

```{r}
library(shiny)
library(reactablePlus)

# The data frame drives the rows — must contain the id column
# and any display columns declared in the row definition.
my_data <- data.frame(
  id   = c("row1", "row2", "row3"),
  name = c("Alice", "Bob", "Carol")
)

rows <- row_def(
  id_col = "id",
  display_cols = list(
    display_col("name", "Name")
  )
)

cols <- list(
  # Simplest form — label == value
  input_col("role", "Role", "dropdown",
            choices = c("Analyst", "Engineer", "Manager")),

  # Named vector — display labels differ from stored values
  input_col("status", "Status", "dropdown",
            choices = c("Active" = "active", "On Leave" = "leave")),

  input_col("score", "Score", "numeric", min = 0, max = 100, step = 1),
  input_col("active", "Active", "checkbox")
)

ui <- fluidPage(
  editable_table_ui("demo")
)

server <- function(input, output, session) {
  editable_table_server(
    "demo",
    data_r   = reactive(my_data),
    row_spec = rows,
    col_spec = cols
  )
}

shinyApp(ui, server)
```

Dropdown choices accept three formats: a plain character vector
(`c("High", "Medium", "Low")`), a named character vector where names
become labels (`c("High" = "high")`), or the full list-of-lists form
when you need non-character values like `list(label = "Good", value = 3)`.
All three are normalized internally via `normalize_choices()`.

Plain lists are still accepted for backward compatibility — the
constructors are the recommended approach for new code.

### 2. Config-driven module (widget-powered)

The higher-level module uses `table_config()` and `widget_col()` to
declaratively describe a table that renders picker widgets in each cell.
It handles mutual exclusion, fill-down, gear toggles, and data marshaling
out of the box.

```{r}
library(shiny)
library(reactablePlus)

cfg <- table_config(
  row_keys   = c("PK", "K", "01", "02"),
  row_labels = c("PreK", "Kindergarten", "1st Grade", "2nd Grade"),
  columns    = list(
    widget_col("school", "school_picker", "School", min_width = 300),
    widget_col("attendance", "attendance_picker", "Attendance", width = 200),
    widget_col("notes", "notes_input", "Notes", min_width = 150)
  )
)

ui <- fluidPage(
  config_table_ui("history", cfg)
)

server <- function(input, output, session) {
  config_table_server("history", cfg,
    data_r    = reactive(NULL),
    search_fn = function(query, grade_key) {
      # Return a data frame of matching schools
      data.frame(name = "Example School", nces_id = "123456")
    }
  )
}

shinyApp(ui, server)
```

## Using picker widgets standalone

You don't have to use the table modules. Every picker widget works as a
standalone Shiny input.

```{r}
ui <- fluidPage(
  useReactablePlus(),
  schoolPickerInput("my_school", grade_key = "PK",
                    grade_label = "PreK"),
  attendancePickerInput("my_att", grade_label = "PreK"),
  notesInput("my_notes", placeholder = "Type here...")
)

server <- function(input, output, session) {
  observeEvent(input$my_school, {
    message("School selected: ", input$my_school$name)
  })
}

shinyApp(ui, server)
```

## Embedding widgets inside reactable

When placing widgets inside `reactable::colDef(cell = ...)`, two things
are required:

1. Call `useReactablePlus()` in the UI to load JS/CSS at the page level
   (reactable's React renderer doesn't process `htmlDependency` objects).

2. Wrap the reactable widget with `bindPickersOnRender()` to initialize
   Shiny input bindings after the table DOM is created.

```{r}
ui <- fluidPage(
  useReactablePlus(),
  reactable::reactableOutput("tbl")
)

server <- function(input, output, session) {
  output$tbl <- reactable::renderReactable({
    tbl <- reactable::reactable(
      data.frame(grade = c("PK", "K"), notes = c("", "")),
      columns = list(
        notes = reactable::colDef(
          html = TRUE,
          cell = function(value, index) {
            as.character(notesInput(
              paste0("note_", index),
              value = value,
              placeholder = "Add a note"
            ))
          }
        )
      )
    )
    bindPickersOnRender(tbl)
  })
}

shinyApp(ui, server)
```

## Next steps

- Browse the [Reference](reference/index.html) for full documentation of
  every function.
- See the [Config-Driven Table Module](reference/config_table_ui.html) docs
  for mutual exclusion, fill-down, and gear toggle configuration.
- See the [Raw-Spec Table Module](reference/editable_table_ui.html) docs
  for gating, selection, and reset behavior.
