---
title: "Getting Started with reactablePlus"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with reactablePlus}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

reactablePlus extends [reactable](https://glin.github.io/reactable/) with
inline editing and popover-based picker widgets for Shiny. Build editable
tables declaratively with `table_config()` and `widget_col()`, wire them
up with `config_table_ui()` / `config_table_server()`, and get a fully
interactive table with gating, row selection, reset, and live data output.

## Installation

```{r}
# install.packages("pak")
pak::pak("matthewcurrier/reactablePlus")
```

## Quick start

A working editable table in under 30 lines:

```{r}
library(shiny)
library(reactablePlus)

cfg <- table_config(
  row_keys   = c("row1", "row2", "row3"),
  row_labels = c("Alice", "Bob", "Carol"),

  columns = list(
    widget_col("role", "dropdown", "Role",
      options = list(choices = c("Analyst", "Engineer", "Manager"))
    ),
    widget_col("score", "numeric", "Score",
      options = list(min = 0, max = 100, step = 1)
    ),
    widget_col("active", "checkbox", "Active"),
    widget_col("notes", "text", "Notes",
      options = list(placeholder = "Optional notes...", max_chars = 200)
    )
  ),

  to_output_fn = function(row_state, row_key) {
    data.frame(
      id     = row_key,
      role   = row_state$role %||% NA_character_,
      score  = row_state$score %||% NA_real_,
      active = isTRUE(row_state$active),
      notes  = row_state$notes %||% "",
      stringsAsFactors = FALSE
    )
  }
)

ui <- fluidPage(
  config_table_ui("demo", cfg)
)

server <- function(input, output, session) {
  result <- config_table_server("demo", cfg)

  observe({
    print(result$get_data())
  })
}

shinyApp(ui, server)
```


## Architecture at a glance

reactablePlus is organized in two layers. Use either independently or
combine them.

**Layer 1 — Picker widgets.** Standalone Shiny inputs
(`schoolPickerInput`, `attendancePickerInput`, `homeschoolPickerInput`,
`notesInput`, `gearPopoverInput`) with custom JS bindings. Use them
anywhere you'd use a regular Shiny input.

**Layer 2 — Config-driven table module.** A Shiny module
(`config_table_ui` / `config_table_server`) that builds an editable
reactable from a `table_config()` object. Supports both primitive
inputs (dropdown, numeric, date, checkbox, toggle, text) and popover
picker widgets — with gating, row selection, reset, mutual exclusion,
and fill-down.


## Primitive input types

`widget_col()` supports six inline HTML input types. Type-specific
options go in the `options` list.

```{r}
columns <- list(
  # Dropdown — choices can be a character vector, named vector, or
  # list of list(label, value). All three are normalized internally.
  widget_col("status", "dropdown", "Status",
    options = list(
      choices = c("Active" = "active", "On Leave" = "leave", "Terminated" = "term")
    )
  ),

  # Numeric — min is required; max and step are optional
  widget_col("hours", "numeric", "Hours",
    options = list(min = 0, max = 80, step = 0.5)
  ),

  # Date — with optional bounds
  widget_col("start", "date", "Start Date",
    options = list(min_date = "2020-01-01", max_date = "2030-12-31")
  ),

  # Checkbox — no extra options
  widget_col("enrolled", "checkbox", "Enrolled"),

  # Toggle — rendered as an on/off button
  widget_col("billable", "toggle", "Billable"),

  # Text — with optional max_chars and placeholder
  widget_col("notes", "text", "Notes",
    options = list(max_chars = 500, placeholder = "Enter notes...")
  )
)
```

## Dropdown choices: three formats

All three are normalized via `normalize_choices()`. Use the simplest
format that fits.

```{r}
# 1. Plain vector — label equals value
widget_col("color", "dropdown", "Color",
  options = list(choices = c("Red", "Blue", "Green"))
)

# 2. Named vector — names become display labels
widget_col("status", "dropdown", "Status",
  options = list(choices = c("Active" = "active", "On Leave" = "leave"))
)

# 3. List of lists — for non-character values
widget_col("rating", "dropdown", "Rating",
  options = list(choices = list(
    list(label = "Good", value = 3),
    list(label = "OK", value = 2),
    list(label = "Poor", value = 1)
  ))
)
```


## Gating (conditional inputs)

Use the `gate` parameter on `widget_col()` to lock a column until
conditions are met. Gates are purely server-driven — no client-side
JavaScript needed.

```{r}
cfg <- table_config(
  row_keys   = c("r1", "r2", "r3"),
  row_labels = c("Item A", "Item B", "Item C"),

  columns = list(
    widget_col("status", "dropdown", "Status",
      options = list(choices = c("Active" = "active", "Inactive" = "inactive"))
    ),
    # Score is locked until Status = "active"
    widget_col("score", "numeric", "Score",
      options = list(min = 0, max = 100),
      gate = list(
        list(type = "value", col_id = "status", values = c("active"))
      )
    )
  ),

  to_output_fn = function(row_state, row_key) {
    data.frame(
      id     = row_key,
      status = row_state$status %||% NA_character_,
      score  = row_state$score %||% NA_real_,
      stringsAsFactors = FALSE
    )
  }
)
```

`table_config()` automatically marks controller columns (here,
`"status"`) as `triggers_rerender = TRUE` so the table re-renders
when they change. Gated values are enforced server-side in
`get_data()` — a locked column's value is forced to its `empty_value`.

Gates can also require row selection:

```{r}
widget_col("priority", "dropdown", "Priority",
  options = list(choices = c("High", "Medium", "Low")),
  gate = list(
    list(type = "selected"),
    list(type = "value", col_id = "status", values = c("active"))
  )
)
```

All conditions use AND logic — every condition must pass for the
input to unlock.


## Row selection and reset

Enable row selection with `selectable = TRUE` and a reset button
with `show_reset = TRUE`:

```{r}
cfg <- table_config(
  row_keys   = c("r1", "r2"),
  row_labels = c("Item A", "Item B"),
  selectable = TRUE,
  show_reset = TRUE,

  columns = list(
    widget_col("notes", "text", "Notes")
  ),

  to_output_fn = function(row_state, row_key) {
    data.frame(
      id       = row_key,
      selected = isTRUE(row_state$.selected),
      notes    = row_state$notes %||% "",
      stringsAsFactors = FALSE
    )
  }
)

# In the server:
result <- config_table_server("demo", cfg)

# result$get_data()     — reactive data frame with gate enforcement
# result$selected_ids() — reactive character vector of selected row keys
```

Reset clears all inputs to their `empty_value` and deselects all rows
in a single reactive flush.


## Picker widgets in tables

For complex inputs — typeahead search, multi-field popovers, rating
scales — use the picker widget types:

```{r}
cfg <- table_config(
  row_keys   = c("PK", "K", "01"),
  row_labels = c("PreK", "K", "1st"),

  columns = list(
    widget_col("school", "school_picker", "School", min_width = 300),
    widget_col("attendance", "attendance_picker", "Attendance", width = 200),
    widget_col("notes", "notes_input", "Notes", min_width = 150)
  ),

  # Show a grade badge column
  badge_col   = "grade",
  badge_label = "Grade",
  badge_render_fn = function(row_key, row_label) {
    css_class <- paste0("grade-badge g-", row_key)
    sprintf('<span class="%s">%s</span>', css_class, row_label)
  }
)
```

Picker widgets require `useReactablePlus()` in the UI (which
`config_table_ui()` includes automatically) and
`bindPickersOnRender()` for binding initialization (which
`config_table_server()` handles internally).


## Using picker widgets standalone

Every picker widget works as a standalone Shiny input outside of
tables. Call `useReactablePlus()` in your UI to load the JS and CSS.

```{r}
library(shiny)
library(reactablePlus)

ui <- fluidPage(
  useReactablePlus(),
  schoolPickerInput("my_school", grade_key = "PK",
                    grade_label = "PreK"),
  attendancePickerInput("my_att", grade_label = "PreK"),
  notesInput("my_notes", placeholder = "Type here...")
)

server <- function(input, output, session) {
  observeEvent(input$my_school, {
    message("School selected: ", input$my_school$name)
  })
}

shinyApp(ui, server)
```


## Embedding widgets inside a custom reactable

When placing widgets inside `reactable::colDef(cell = ...)` outside
of `config_table`, two extra steps are needed because reactable's
React-based rendering doesn't process Shiny input bindings
automatically.

1. Call `useReactablePlus()` in the UI to load JS/CSS at the page level.
2. Wrap the reactable widget with `bindPickersOnRender()` to initialize
   input bindings after the table DOM is created.

```{r}
library(shiny)
library(reactablePlus)

ui <- fluidPage(
  useReactablePlus(),
  reactable::reactableOutput("tbl")
)

server <- function(input, output, session) {
  output$tbl <- reactable::renderReactable({
    tbl <- reactable::reactable(
      data.frame(grade = c("PK", "K"), notes = c("", "")),
      columns = list(
        notes = reactable::colDef(
          html = TRUE,
          cell = function(value, index) {
            as.character(notesInput(
              paste0("note_", index),
              value = value,
              placeholder = "Add a note"
            ))
          }
        )
      )
    )
    bindPickersOnRender(tbl)
  })
}

shinyApp(ui, server)
```


## Configurable hooks

`table_config()` provides hooks that keep the core module generic
while letting domain-specific apps customize appearance:

- `badge_col` / `badge_label` / `badge_render_fn` — opt-in label
  column with custom rendering (e.g., grade badges with CSS classes)
- `row_class_fn(row_key, row_state)` — custom CSS classes per row
- `to_output_fn(row_state, row_key)` — marshal row state into a
  data frame for your output schema
- `from_saved_fn(db_row, col_specs)` — parse saved data into row state
- `toolbar_stats_fn(rows, row_keys)` — dynamic toolbar content
- `year_col` / `year_range` — opt-in editable year spinner column


## Grade utilities

For PreK–12 education apps, reactablePlus exports helper functions
in `grade-utils.R`:

- `gradeChoices()` — named character vector of the 14-grade ladder
- `gradeLabelMap()` — grade key → display label lookup
- `gradeIndex()` — grade key → numeric index for ordering
- `gradeInRange()` — check if a grade falls within a school's range

These are convenience exports, not required by the table infrastructure.


## Migration from the raw-spec API

The original `editable_table_ui()` / `editable_table_server()` module
with `row_def()`, `display_col()`, and `input_col()` constructors is
now deprecated. The config-driven API (`config_table_*`) supports all
the same input types plus picker widgets, gating, selection, and reset
with a simpler state model.

Quick mapping:

| Raw-spec | Config-driven |
|---|---|
| `row_def(id_col, display_cols)` | `table_config(row_keys, row_labels)` |
| `display_col("name", "Name")` | included in `row_labels` |
| `input_col("x", "X", "dropdown", choices = ...)` | `widget_col("x", "dropdown", "X", options = list(choices = ...))` |
| `editable_table_ui("id")` | `config_table_ui("id", cfg)` |
| `editable_table_server("id", data_r, row_spec, col_spec)` | `config_table_server("id", cfg)` |
| `result$current_data` | `result$get_data()` |
| `result$selected_ids` | `result$selected_ids()` |

The deprecated functions still work and emit a warning pointing to
the replacement. They will be removed in a future major release.


## Next steps

- Browse the [Reference](reference/index.html) for full documentation
  of every exported function.
- See `table_config()` for mutual exclusion, fill-down, and gear
  toggle configuration.
- See `widget_col()` for the full list of types, options, and gate
  conditions.
