---
title: "Getting Started with reactablePlus"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

reactablePlus extends [reactable](https://glin.github.io/reactable/) with
inline editing widgets and two Shiny module layers for building editable
tables. It provides everything from low-level input constructors to a
fully declarative, config-driven table module with picker widgets,
mutual exclusion, fill-down, and gear toggles.

## Installation

```{r}
# install.packages("pak")
pak::pak("matthewcurrier/reactablePlus")
```

## Architecture at a glance

reactablePlus is organized in three layers. You can use any layer
independently or combine them.

**Layer 1 — Picker widgets.** Standalone Shiny inputs
(`schoolPickerInput`, `attendancePickerInput`, `homeschoolPickerInput`,
`notesInput`, `gearPopoverInput`) with custom JS bindings. Use them
anywhere you'd use a regular Shiny input.

**Layer 2 — Raw-spec table module.** A Shiny module
(`editable_table_ui` / `editable_table_server`) that renders an editable
reactable from declarative row and column definitions. Supports dropdowns,
numeric spinners, checkboxes, toggles, date pickers, and text fields —
with gating, row selection, and reset.

**Layer 3 — Config-driven table module.** A higher-level Shiny module
(`config_table_ui` / `config_table_server`) that renders picker widgets
inside table cells from a `table_config()` object. Handles mutual
exclusion, fill-down, gear toggles, and data marshaling out of the box.


## Layer 2: The raw-spec table module

This is the most flexible layer. You define rows and columns with
constructor functions, pass in a data frame, and get an interactive
editable table.

### Defining rows with `row_def()`

A row definition tells the module which column in your data frame
identifies rows, which columns to display as read-only, and whether
rows are selectable.

```{r}
rows <- row_def(
  id_col = "id",
  display_cols = list(
    display_col("name", "Name"),
    display_col("department", "Department")
  ),
  selectable = TRUE
)
```

### Defining input columns with `input_col()`

Each `input_col()` describes one editable column. The `type` argument
determines what kind of input renders in each cell.

```{r}
cols <- list(
  input_col("role", "Role", "dropdown",
            choices = c("Analyst", "Engineer", "Manager")),
  input_col("score", "Score", "numeric",
            min = 0, max = 100, step = 1),
  input_col("active", "Active", "checkbox"),
  input_col("start_date", "Start Date", "date",
            min_date = "2020-01-01"),
  input_col("notes", "Notes", "text", max_chars = 200)
)
```

The constructors validate your configuration immediately — a missing
`min` on a numeric column or a bad `choices` format on a dropdown will
error at definition time, not when the app renders.

### Dropdown choices: three formats

`input_col()` accepts dropdown choices in three formats, from simplest
to most flexible.

```{r}
# 1. Plain vector — label equals value
input_col("color", "Color", "dropdown",
          choices = c("Red", "Blue", "Green"))

# 2. Named vector — names become display labels
input_col("status", "Status", "dropdown",
          choices = c("Active" = "active", "On Leave" = "leave"))

# 3. List of lists — for non-character values
input_col("rating", "Rating", "dropdown",
          choices = list(
            list(label = "Good", value = 3),
            list(label = "OK", value = 2),
            list(label = "Poor", value = 1)
          ))
```

All three are normalized internally via `normalize_choices()`. Use
the simplest format that fits your use case.

### Putting it together

```{r}
library(shiny)
library(reactablePlus)

my_data <- data.frame(
  id   = c("row1", "row2", "row3"),
  name = c("Alice", "Bob", "Carol")
)

rows <- row_def(
  id_col = "id",
  display_cols = list(display_col("name", "Name"))
)

cols <- list(
  input_col("role", "Role", "dropdown",
            choices = c("Analyst", "Engineer", "Manager")),
  input_col("score", "Score", "numeric", min = 0, max = 100, step = 1),
  input_col("active", "Active", "checkbox")
)

ui <- fluidPage(
  editable_table_ui("demo")
)

server <- function(input, output, session) {
  result <- editable_table_server(
    "demo",
    data_r   = reactive(my_data),
    row_spec = rows,
    col_spec = cols
  )

  # result$current_data is a reactive tibble of all input values
  # result$selected_ids is a reactive vector of checked row IDs
}

shinyApp(ui, server)
```

### A realistic example with mtcars

```{r}
library(shiny)
library(reactablePlus)

mtcars$id <- rownames(mtcars[1:20,])
rownames(mtcars) <- NULL

rows <- row_def(
  id_col = "id",
  display_cols = list(
    display_col("cyl", "Cylinders"),
    display_col("qsec", "1/4 Mile Time")
  )
)

cols <- list(
  input_col("disp_cat", "Displacement", "dropdown",
            choices = c("High" = "high", "Medium" = "medium", "Low" = "low")),
  input_col("score", "Score", "numeric", min = 0, max = 100, step = 1),
  input_col("flagged", "Flagged", "checkbox"),
  input_col("assessment", "Assessment", "dropdown",
            choices = list(
              list(label = "Good", value = 3),
              list(label = "OK", value = 2),
              list(label = "Poor", value = 1)
            ))
)

ui <- fluidPage(
  editable_table_ui("cars")
)

server <- function(input, output, session) {
  editable_table_server(
    "cars",
    data_r   = reactive(mtcars),
    row_spec = rows,
    col_spec = cols
  )
}

shinyApp(ui, server)
```


## Layer 3: The config-driven table module

The higher-level module uses `table_config()` and `widget_col()` to
declaratively describe a table that renders picker widgets in each cell.
This is best suited for domain-specific applications where you need
popover-based editors, school search, attendance tracking, or similar
complex cell interactions.

```{r}
library(shiny)
library(reactablePlus)

cfg <- table_config(
  row_keys   = c("PK", "K", "01", "02"),
  row_labels = c("PreK", "Kindergarten", "1st Grade", "2nd Grade"),
  columns    = list(
    widget_col("school", "school_picker", "School", min_width = 300),
    widget_col("attendance", "attendance_picker", "Attendance", width = 200),
    widget_col("notes", "notes_input", "Notes", min_width = 150)
  )
)

ui <- fluidPage(
  config_table_ui("history", cfg)
)

server <- function(input, output, session) {
  config_table_server("history", cfg,
    data_r    = reactive(NULL),
    search_fn = function(query, grade_key) {
      # Return a data frame of matching schools
      data.frame(name = "Example School", nces_id = "123456")
    }
  )
}

shinyApp(ui, server)
```


## Layer 1: Using picker widgets standalone

Every picker widget works as a standalone Shiny input — you don't need
the table modules. Just call `useReactablePlus()` in your UI to load
the required JS and CSS.

```{r}
library(shiny)
library(reactablePlus)

ui <- fluidPage(
  useReactablePlus(),
  schoolPickerInput("my_school", grade_key = "PK",
                    grade_label = "PreK"),
  attendancePickerInput("my_att", grade_label = "PreK"),
  notesInput("my_notes", placeholder = "Type here...")
)

server <- function(input, output, session) {
  observeEvent(input$my_school, {
    message("School selected: ", input$my_school$name)
  })
}

shinyApp(ui, server)
```


## Embedding widgets inside reactable cells

When placing widgets inside `reactable::colDef(cell = ...)`, two extra
steps are required because reactable's React-based rendering doesn't
process Shiny's `htmlDependency` objects or initialize input bindings
automatically.

1. Call `useReactablePlus()` in the UI to load JS/CSS at the page level.
2. Wrap the reactable widget with `bindPickersOnRender()` to initialize
   Shiny input bindings after the table DOM is created.

```{r}
library(shiny)
library(reactablePlus)

ui <- fluidPage(
  useReactablePlus(),
  reactable::reactableOutput("tbl")
)

server <- function(input, output, session) {
  output$tbl <- reactable::renderReactable({
    tbl <- reactable::reactable(
      data.frame(grade = c("PK", "K"), notes = c("", "")),
      columns = list(
        notes = reactable::colDef(
          html = TRUE,
          cell = function(value, index) {
            as.character(notesInput(
              paste0("note_", index),
              value = value,
              placeholder = "Add a note"
            ))
          }
        )
      )
    )
    bindPickersOnRender(tbl)
  })
}

shinyApp(ui, server)
```


## Choosing the right layer

| Need | Use |
|---|---|
| A quick editable table with standard inputs | Layer 2: `row_def()` + `input_col()` + `editable_table_server()` |
| Popover-based pickers with cross-column logic | Layer 3: `table_config()` + `widget_col()` + `config_table_server()` |
| A single picker input outside a table | Layer 1: `schoolPickerInput()`, `notesInput()`, etc. |
| Widgets inside a custom reactable layout | Layer 1 + `useReactablePlus()` + `bindPickersOnRender()` |


## Next steps

- Browse the [Reference](reference/index.html) for full documentation of
  every exported function.
- See `config_table_ui()` docs for mutual exclusion, fill-down, and
  gear toggle configuration.
- See `editable_table_ui()` docs for gating, selection, and reset
  behavior.
- Use `validate_col_spec()` and `validate_row_spec()` to debug
  configuration issues — they produce specific, actionable error messages.

